Density-dependent selection in evolutionary genetics: a lottery model of Grime's triangle

### 2017-04-26 manuscript.tex
latest

We assume that reproductively mature individuals (``adults'') each require their own territory to survive and reproduce (Fig.~\ref{fig:lottery}). All territories are identical, and the total number of territories is $T$. Time $t$ advances in discrete iterations, each representing the time from birth to reproductive maturity. In iteration $t$, the number of adults of the $i$'th genotype is $n_i(t)$, the total number of adults is $N(t)=\sum_i n_i(t)$, and the number of unoccupied territories is $U(t)=T-N(t)$. We assume that the $n_i$ and $T$ are large enough that stochastic fluctuations in the $n_i$ (``drift'') can be ignored. We derive deterministic equations for the expected change in the $n_i$ over time, leaving the evaluation of drift for future work.

@@

When a new mutant genotype $j$ appears, it starts with one individual $n_j=1$. While the mutant lineage remains at low-abundance, $n_j$ will behave stochastically, but if its expected growth rate is positive and $n_j$ becomes large enough, $n_j$ will effectively grow deterministically according to Eq. \eqref{eq:master}. This transition occurs at an abundance of order $1/r$ \citep{uecker_2011}, where $r=\Delta n_i/n_i$ is the mutant lineage's growth rate. Since here we do not evaluate drift in the extended lottery model, we do not attempt to calculate the probability that mutants escape the initial stochastic phase, and restrict our attention to the earliest deterministic behavior of rare genotypes while they are still at negligibly low frequency. We simply invoke the well known result this probability is proportional to $r$, with a proportionality factor typically of order one (Haldane's formula; \citealt{uecker_2011}). The fixation of neutral mutations is exceedingly unlikely (probability of order $1/N$). Consequently, the direction of evolutionary change is determined by the mutational trait changes which are available and also confer an appreciable fitness benefit, where availability is subject to constraints imposed by the environment.

>>>>

We assume that reproductively mature individuals (``adults'') each require their own territory to survive and reproduce (Fig.~\ref{fig:lottery}). All territories are identical, and the total number of territories is $T$. Time $t$ advances in discrete iterations, each representing the time from birth to reproductive maturity. In iteration $t$, the number of adults of the $i$'th genotype is $n_i(t)$, the total number of adults is $N(t)=\sum_i n_i(t)$, and the number of unoccupied territories is $U(t)=T-N(t)$.

We assume that the $n_i$ and $T$ are large enough that stochastic fluctuations in the $n_i$ (``drift'') can be ignored. We derive deterministic equations for the expected change in the $n_i$ over time, leaving the evaluation of drift for future work. In particular, we do not evaluate the initial stochastic behaviour of adaptive mutant lineages while they are at low abundances. We restrict our attention to the earliest deterministic behavior of mutant lineages while they are still at low frequency (the transition to deterministic growth occurs at an abundance of order equal to their inverse expected absolute growth rate; \citealt{uecker_2011}).

@@

We simply invoke the well known result this probability is proportional to $r$, with a proportionality factor typically of order one (Haldane's formula; \citealt{uecker_2011}). The fixation of neutral mutations is exceedingly unlikely (probability of order $1/N$). Consequently, the direction of evolutionary change is determined by the mutational trait changes which are available and also confer an appreciable fitness benefit, where availability is subject to constraints imposed by the environment.
